Fit by design: Developing substrate-specific seed mixtures for functional dike grasslands

### 2023-03-02 manuscript.Rmd
Update manuscript.Rmd

1.  Sowing is a well-established restoration technique to overcome dispersal limitation and site-specific seed mixtures are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, reduce mowing efforts, and generate biodiverse grasslands.
2.  A factorial field experiment addressed this topic on a dike at the Danube River in SE Germany in the 2018--2021 time period. On 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north or south exposition, and measured the persistence of sown species, the Favourable Conservation Status (FCS), and the successional distance to reference sites as calculated from an ordination.
3.  Sand admixture had a slightly positive effect on target variables, while substrate depth, seed density and mixture type had no effects on species persistence or FCS.
4.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry grasslands.
5.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment and the addition and removal of a geotextile. This initial negative effect remained over the entire observation period.
6.  Synthesis and applications: Site-adapted seed mixtures make restoration more efficient. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt and re-sowing after unfavourable conditions.

@@

We could identify a statistically clear positive effect of the sand admixture on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:persistence), \@ref(fig:fcs), \@ref(fig:recovery)). For all three response variables, the vegetation developed strongly positive after one year, but the recovery rate slowed down in the following years. Both expositions had similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:persistence)). The interactions of restoration treatments were neither clear nor strong (Figures \@ref(fig:persistence), \@ref(fig:fcs), \@ref(fig:recovery)). Persistence of both seed mixture types was slightly positively affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:persistence)).

@@

The reference sites had a larger variation than the seed mixtures and were close to the seed mixtures but hardly overlapped (Figure \@ref(fig:nmds)). The positions of the reference sites were shifted to the left in comparison to the seed mixtures, which means in the direction of early successional states. Nonetheless, they still differed from the negative references of ruderal vegetation. Negative references were only available on the south exposition and they were located in the NMDS between the positive reference sites and the state of restored plots in 2021. Hay meadow-seed mixtures led to a closer development to hay meadow references than dry grasslands to their references (Figure \@ref(fig:effects), \@ref(fig:recovery)). This was especially the case in north exposition (Figure \@ref(fig:nmds)). \clearpage \newpage

@@

@@

The seed mixtures and their positive reference sites were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the NMDS suggests that the seed mixture represents a later successional stage than the references. The NMDS shows a slightly better adaptation of hay meadows to the north exposition than of dry grasslands (Figure \@ref(fig:recovery)). This can be expected from the requirements of hay meadows for mesic conditions which can be provided on north-exposed dike slopes [@Batori.2020; @Oberdorfer.1993]. On the southern exposition, the plots of hay meadows developed rather to dry grassland references which indicates an inefficient restoration by using a non-adapted seed mixture.

@@

```{r tab.id = "substrate", ft.align = "left", tab.cap = "Substrates"}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  select(-starts_with("lfl")) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      select(-starts_with("lfl")) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(
    i = 1, j = 10, part = "header",
    value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))
    ) %>%
  compose(
    i = 1, j = 11, part = "header",
    value = as_paragraph("K", as_sub("2"), "O")
    ) %>%
  compose(
    i = 1, j = 12, part = "header",
    value = as_paragraph("Mg", as_sup("2+"))
    ) %>%
  compose(
    i = 1, j = 14, part = "header",
    value = as_paragraph("CaCO", as_sub("3"))
    ) %>%
  compose(
    i = 2, j = 10, part = "header",
    value = as_paragraph("mg 100 g", as_sup("−1"))
    ) %>%
  compose(
    i = 2, j = 11, part = "header",
    value = as_paragraph("mg 100 g", as_sup("−1"))
    ) %>%
  compose(
    i = 2, j = 12, part = "header",
    value = as_paragraph("mg 100 g", as_sup("−1"))
    ) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

```{r fig.id = "effects", fig.asp = 0.3030, fig.width = 6.49, fig.cap = "The posterior density distributions (grey) of the treatments calculated over all four surveyed years of sown grassland communities at a river dike. Furthermore, the median and the 66% and 95% credible intervals are shown: the persistence of sown species (A), the Favourable Conservation Status (FCS; B) and the successional distance to reference (recovery; C). The FCS is the ratio of target species to non-target species. Note that the zero lines indicate that both levels have equal values. This means, e.g., that hay meadows are closer to their reference than dry grasslands  (C)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_2_800dpi_16.5x5cm.tiff")
  )
```

@@

```{r fig.id = "persistence", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "Persistence (1-*B*~sor~) of sown species of sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with two different seed mixture types. Twenty species were sown per plot. Shown are the medians and 95% credible intervals of the posterior distributions, which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here(
    "outputs", "figures", "figure_3_persistence_epred_800dpi_16.5x8cm.tiff"
    )
  )
```

@@

```{r fig.id = "fcs", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The Favourable Conservation Status (FCS) of sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with two different seed mixtures. The zero line indicates that target and non-target species are balanced. Positive values indicate that there are more target species. Shown are the medians and 95% credible intervals of the posterior distributions, which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_4_fcs_epred_800dpi_16.5x8cm.tiff")
  )
```

## Figure 5 {.unnumbered}

```{r fig.id = "recovery", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The recovery or successional distance (*d~jt,0~*) of sown grassland communities at a river dike in comparison to the respective reference sites (mesic hay meadows or semi-dry calcareous grasslands). The plots had substrates with different sand admixtures and were sown with two different seed mixtures. The zero lines indicate the mean position of the reference sites for each habitat type on the NMDS axis 1 (Figure 7). The grey area marks the standard deviation of the position of the reference sites. Shown are the medians and 95% credible intervals of the posterior distributions, which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here(
    "outputs", "figures", "figure_5_recovery_epred_800dpi_16.5x8cm.tiff"
    )
  )
```

## Figure 6 {.unnumbered}

```{r fig.id = "nmds", fig.asp = 0.9696, fig.width = 6.49, fig.cap = ""}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_6_nmds_800dpi_16.5x16cm.tiff")
  )
```

>>>>

1.  Sowing is a well-established restoration technique to overcome dispersal limitation and site-specific seed mixtures are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, reduce mowing efforts, and generate biodiverse grasslands.
2.  A factorial field experiment addressed this topic on a dike at the Danube River in SE Germany in the 2018--2021 time period. On 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north or south exposition, and measured the persistence of sown species, the Favourable Conservation Status (FCS), and the successional distance to reference sites as calculated from an ordination.
3.  Sand admixture had a slightly positive effect on target variables, while substrate depth, seed density and mixture type had no effects on species persistence or FCS.
4.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry grasslands.
5.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment and the addition and removal of a geotextile. This initial negative effect remained over the entire observation period.
6.  Synthesis and applications: Site-adapted seed mixtures make restoration more efficient. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt and re-sowing after unfavourable conditions. \clearpage \newpage

@@

@@

The reference sites had a larger variation than the seed mixtures and were close to the seed mixtures but hardly overlapped (Figure \@ref(fig:nmds)). The positions of the reference sites were shifted to the left in comparison to the seed mixtures, which means in the direction of early successional states. Nonetheless, they still differed from the negative references of ruderal vegetation. Negative references were only available on the south exposition and they were located in the NMDS between the positive reference sites and the state of restored plots in 2021. Hay meadow-seed mixtures led to a closer development to hay meadow references than dry grasslands to their references (Figure \@ref(fig:effects)A, \@ref(fig:interactions)A). This was especially the case in north exposition (Figure \@ref(fig:nmds)).

We could identify a statistically clear positive effect of the sand admixture on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:interactions)). For all three response variables, the vegetation developed strongly positive after one year, but the recovery rate slowed down in the following years. Both expositions had similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:interactions)B). The interactions of restoration treatments were neither clear nor strong (Figures \@ref(fig:interactions)). Persistence of both seed mixture types was slightly positively affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:interactions)B). \clearpage \newpage

@@

The seed mixtures and their positive reference sites were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the NMDS suggests that the seed mixture represents a later successional stage than the references. The NMDS shows a slightly better adaptation of hay meadows to the north exposition than of dry grasslands (Figure \@ref(fig:interactions)A). This can be expected from the requirements of hay meadows for mesic conditions which can be provided on north-exposed dike slopes [@Batori.2020; @Oberdorfer.1993]. On the southern exposition, the plots of hay meadows developed rather to dry grassland references which indicates an inefficient restoration by using a non-adapted seed mixture.

@@

@@

```{r tab.id = "substrate", ft.align = "left", tab.cap = ""}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  select(-starts_with("lfl")) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      select(-starts_with("lfl")) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(
    i = 1, j = 10, part = "header",
    value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))
    ) %>%
  compose(
    i = 1, j = 11, part = "header",
    value = as_paragraph("K", as_sub("2"), "O")
    ) %>%
  compose(
    i = 1, j = 12, part = "header",
    value = as_paragraph("Mg", as_sup("2+"))
    ) %>%
  compose(
    i = 1, j = 14, part = "header",
    value = as_paragraph("CaCO", as_sub("3"))
    ) %>%
  compose(
    i = 2, j = 10, part = "header",
    value = as_paragraph("mg 100 g", as_sup("−1"))
    ) %>%
  compose(
    i = 2, j = 11, part = "header",
    value = as_paragraph("mg 100 g", as_sup("−1"))
    ) %>%
  compose(
    i = 2, j = 12, part = "header",
    value = as_paragraph("mg 100 g", as_sup("−1"))
    ) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

```{r fig.id = "nmds", fig.asp = 0.9696, fig.width = 6.49, fig.cap = ""}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_2_nmds_800dpi_16.5x16cm.tiff")
  )
```

@@

```{r fig.id = "effects", fig.asp = 0.3030, fig.width = 6.49, fig.cap = ""}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_3_800dpi_16.5x5cm.tiff")
  )
```

Effects of treatments on the development of sown grassland communities at a river dike. The posterior density distributions (grey) is calculated over all four surveyed years and both expositions. Shown are the medians, 66% and 95% credible intervals, which were derived from a Bayesian linear mixed-effects model (BLMM). (A) The persistence of sown species, (B) the Favourable Conservation Status (FCS) and (C) the recovery completeness compared to reference sites. The FCS is the ratio of target species to non-target species. Note that the zero lines indicate that both levels have equal values. This means, e.g., that hay meadows are closer to their reference than dry grasslands (C).

@@

```{r fig.id = "interactions", fig.asp = 1.2121, fig.width = 6.49, fig.cap = ""}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_4_800dpi_16.5x20cm.tiff")
  )
```

The development sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with two different seed mixture types. Three indices are evaluated. (A) Recovery completeness (*d~jt~*~,0~): the zero lines indicate the mean position of the reference sites for each habitat type on the NMDS axis 1 (Figure \@ref(fig:nmds)). The grey area marks the standard deviation of the position of the reference sites. (B) Persistence of sown species: losses component of the temporal beta-diversity index (1 − *B*~sor~) (C) Favourable Conservation Status (FCS): the zero line indicates that target and non-target species are balanced. Positive values indicate that there are more target species. Shown are the medians and 95% credible intervals of the posterior distributions, which were derived from a Bayesian linear mixed-effects model (BLMM).
